---
title: "Analysis"
format:
  html:
    code-tools: true
    code-fold: true
---

# Exploratory Data Analysis

## **Suicides by Gender**

We used the bar plot method to compare suicides by gender regardless of age and reason. You can find the graph and the code we used below.

```{r}
#| output: false
library(tidyverse) 
library(dslabs)
library(readxl)
data1 <- read_excel("C:/Users/kubil/Desktop/veri seti.xls")
save(data1, file = "semicolon.RData")
```

```{r}
#| code-fold: true
filtered_data_male <- filter(data1, Sex == "Male", `Age group`=="Total")
male = sum(as.numeric(filtered_data_male$Total))
filtered_data_female <- filter(data1, Sex == "Female", `Age group`=="Total")
female = sum(as.numeric(filtered_data_female$Total))
data <- c(male,female)
barplot(data, names.arg = c("Male","Female"), col = "skyblue", main = "Number of Suicides by Gender", xlab = "Sex", ylab = "Total")
```

-From this graph, we can observe that **men are more prone to suicide compared to women**.

## **Suicides by Age Groups**

We used the bar plot method to compare suicides by age groups regardless of gender and reason. You can find the graph and the code we used below.

```{r}
#| code-fold: true
filtered_data_1 <- filter(data1,`Age group`=="<15", Sex == "Total")
filtered_data_2 <- filter(data1,`Age group`=="15-19", Sex == "Total")
filtered_data_3 <- filter(data1,`Age group`=="20-24", Sex == "Total")
filtered_data_4 <- filter(data1,`Age group`=="25-29", Sex == "Total")
filtered_data_5 <- filter(data1,`Age group`=="30-34", Sex == "Total")
filtered_data_6 <- filter(data1,`Age group`=="35-39", Sex == "Total")
filtered_data_7 <- filter(data1,`Age group`=="40-44", Sex == "Total")
filtered_data_8 <- filter(data1,`Age group`=="45-49", Sex == "Total")
filtered_data_9 <- filter(data1,`Age group`=="50-54", Sex == "Total")
filtered_data_10 <- filter(data1,`Age group`=="55-59", Sex == "Total")
filtered_data_11 <- filter(data1,`Age group`=="60-64", Sex == "Total")
filtered_data_12 <- filter(data1,`Age group`=="65-69", Sex == "Total")
filtered_data_13 <- filter(data1,`Age group`=="75 +", Sex == "Total")
age1 = sum(as.numeric(filtered_data_1$Total))
age2 = sum(as.numeric(filtered_data_2$Total))
age3 = sum(as.numeric(filtered_data_3$Total))
age4 = sum(as.numeric(filtered_data_4$Total))
age5 = sum(as.numeric(filtered_data_5$Total))
age6 = sum(as.numeric(filtered_data_6$Total))
age7 = sum(as.numeric(filtered_data_7$Total))
age8 = sum(as.numeric(filtered_data_8$Total))
age9 = sum(as.numeric(filtered_data_9$Total))
age10 = sum(as.numeric(filtered_data_10$Total))
age11 = sum(as.numeric(filtered_data_11$Total))
age12 = sum(as.numeric(filtered_data_12$Total))
age13 = sum(as.numeric(filtered_data_13$Total))
data = c(age1,age2,age3,age4,age5,age6,age7,age8,age9,age10,age11,age12,age13)
barplot(data, names.arg = c("<15","15-19","20-24","25-29","30-34","35-39","40-44","45-49","50-54","55-59","60-64","65-69","75 +"), col = "red", main = "Number of Suicides by Age Group", xlab = "Age Group", ylab = "Total")
```

When looking at this graph, we observe that **suicides are most common among young adults** aged 20-24. We can say that as age progresses beyond 25, suicide rates tend to decrease. Another interesting result is that the age groups with the **lowest suicide rates are children and the elderly.**

```{r}
#| code-fold: true
#| output: false
ffiltered_data_1 <- filter(data1,Year=="2002",`Age group`=="Total", Sex == "Total")
ffiltered_data_2 <- filter(data1,Year=="2003",`Age group`=="Total", Sex == "Total")
ffiltered_data_3 <- filter(data1,Year=="2004",`Age group`=="Total", Sex == "Total")
ffiltered_data_4 <- filter(data1,Year=="2005",`Age group`=="Total", Sex == "Total")
ffiltered_data_5 <- filter(data1,Year=="2006",`Age group`=="Total", Sex == "Total")
ffiltered_data_6 <- filter(data1,Year=="2007",`Age group`=="Total", Sex == "Total")
ffiltered_data_7 <- filter(data1,Year=="2008",`Age group`=="Total", Sex == "Total")
ffiltered_data_8 <- filter(data1,Year=="2009",`Age group`=="Total", Sex == "Total")
ffiltered_data_9 <- filter(data1,Year=="2010",`Age group`=="Total", Sex == "Total")
ffiltered_data_10 <- filter(data1,Year=="2011",`Age group`=="Total", Sex == "Total")
ffiltered_data_11 <- filter(data1,Year=="2012",`Age group`=="Total", Sex == "Total")
ffiltered_data_12 <- filter(data1,Year=="2013",`Age group`=="Total", Sex == "Total")
ffiltered_data_13 <- filter(data1,Year=="2014",`Age group`=="Total", Sex == "Total")
ffiltered_data_14 <- filter(data1,Year=="2015",`Age group`=="Total", Sex == "Total")
ffiltered_data_15 <- filter(data1,Year=="2016",`Age group`=="Total", Sex == "Total")
ffiltered_data_16 <- filter(data1,Year=="2017",`Age group`=="Total", Sex == "Total")
ffiltered_data_17 <- filter(data1,Year=="2018",`Age group`=="Total", Sex == "Total")
ffiltered_data_18 <- filter(data1,Year=="2019",`Age group`=="Total", Sex == "Total")
ffiltered_data_19 <- filter(data1,Year=="2020",`Age group`=="Total", Sex == "Total")
ffiltered_data_20 <- filter(data1,Year=="2021",`Age group`=="Total", Sex == "Total")
ffiltered_data_21 <- filter(data1,Year=="2022",`Age group`=="Total", Sex == "Total")

year1 = sum(as.numeric(ffiltered_data_1$Total))
year2 = sum(as.numeric(ffiltered_data_2$Total))
year3 = sum(as.numeric(ffiltered_data_3$Total))
year4 = sum(as.numeric(ffiltered_data_4$Total))
year5 = sum(as.numeric(ffiltered_data_5$Total))
year6 = sum(as.numeric(ffiltered_data_6$Total))
year7 = sum(as.numeric(ffiltered_data_7$Total))
year8 = sum(as.numeric(ffiltered_data_8$Total))
year9 = sum(as.numeric(ffiltered_data_9$Total))
year10 = sum(as.numeric(ffiltered_data_10$Total))
year11 = sum(as.numeric(ffiltered_data_11$Total))
year12 = sum(as.numeric(ffiltered_data_12$Total))
year13 = sum(as.numeric(ffiltered_data_13$Total))
year14 = sum(as.numeric(ffiltered_data_14$Total))
year15 = sum(as.numeric(ffiltered_data_15$Total))
year16 = sum(as.numeric(ffiltered_data_16$Total))
year17 = sum(as.numeric(ffiltered_data_17$Total))
year18 = sum(as.numeric(ffiltered_data_18$Total))
year19 = sum(as.numeric(ffiltered_data_19$Total))
year20 = sum(as.numeric(ffiltered_data_20$Total))
year21 = sum(as.numeric(ffiltered_data_21$Total))
```

```{r}
#| code-fold: true
data = c(year1,year2,year3,year4,year5,year6,year7,year8,year9,year10,year11,year12,year13,year14,year15,year16,year17,year18,year19,year20,year21)
barplot(data, names.arg = c("2002","2003","2004","2005","2006","2007","2008","2009","2010","2011","2012","2013","2014","2015","2016","2017","2018","2019","2020","2021","2022"), col = "yellow", main = "Number of Suicides by Year", xlab = "Year", ylab = "Total")
```

```{r}
#| code-fold: true
#| output: false
fffiltered_data_1 <- filter(data1,`Age group`=="Total", Sex == "Total")
cause1 = sum(as.numeric(fffiltered_data_1$Illness))
cause2 = sum(as.numeric(fffiltered_data_1$`Family incompatibility`))
cause3 = sum(as.numeric(fffiltered_data_1$`Economic problems`))
cause4 = sum(as.numeric(fffiltered_data_1$`Business failure`))
cause5 = sum(as.numeric(fffiltered_data_1$`Emotional relationship and not marrying the person wanted`))
cause6 = sum(as.numeric(fffiltered_data_1$`Educational failure`))
cause7 = sum(as.numeric(fffiltered_data_1$Other))
unknown <- ifelse(is.na(as.numeric(fffiltered_data_1$Unknown)),0,as.numeric(fffiltered_data_1$Unknown))
cause8 = sum(unknown)
data = c(cause1,cause2,cause3,cause4,cause5,cause6,cause7,cause8)
```

```{r}
#| code-fold: true
barplot(data, names.arg = c("Illness","Family","Economy","Business","Emotional","Educational","Other","Unknown"), col = "brown", main = "Number of Suicides by Reason", xlab = "Suicide Reason", ylab = "Total",cex.names=0.72)
```
