---
title: "Analysis"
format:
  html:
    code-tools: true
    code-fold: true
---

# Exploratory Data Analysis

## Loading Libraries and Datasets

```{r}
#| output: false
library(tidyverse) 
library(dslabs)
library(readxl)
library(dplyr)
library(ggplot2)
library(viridis)
library(hrbrthemes)
data <- read_excel("C:/Users/kubil/Desktop/intihar.xls")
save(data, file = "semicolon1.RData")
nüfus <- read_excel("C:/Users/kubil/Desktop/nüfus.xlsx")
save(nüfus, file = "semicolon2.RData")
```

## Finding the Suicide Rate Using Suicide and Population Datasets

We have data on the number of suicides between 2002 and 2022, but our population data is limited between 2007 and 2022. Since it is more logical to use suicide rate data, we will use data between 2007 and 2022.

```{r}
data <- filter(data,Year >= 2007)
colnames(data) <- c("Year","Age","Sex","Total","Illness","Family","Economy","Business","Emotional","Education","Other","Unknown")
nüfus <- filter(nüfus,Yıl >= 2007,Düzey == "TÜRKİYE")
a = 0
popu = sum(as.numeric(nüfus$Toplam))/length(nüfus$Toplam)
for (i in 1:16){
  data1 <- filter(data, Year == 2023-i)
  data1$Total <- data1$Total/nüfus$Toplam[i]*popu
  data1$Illness <- data1$Illness/nüfus$Toplam[i]*popu
  data1$Family <- data1$Family/nüfus$Toplam[i]*popu
  data1$Economy <- data1$Economy/nüfus$Toplam[i]*popu
  data1$Business <- data1$Business/nüfus$Toplam[i]*popu
  data1$Emotional <- data1$Emotional/nüfus$Toplam[i]*popu
  data1$Education <- data1$Education/nüfus$Toplam[i]*popu
  data1$Other <- data1$Other/nüfus$Toplam[i]*popu
  data1$Unknown <- data1$Unknown/nüfus$Toplam[i]*popu
  a <- rbind(a,data1,deparse.level = 0)
}
data = filter(a,Year != 0)
```

## **Suicides by Gender**

We used the bar plot method to compare suicides by gender regardless of age and reason. You can find the graph and the code we used below.

```{r}
#| code-fold: true
filtered_data_male <- filter(data, Sex == "Male", Age=="Total")
male = sum(as.numeric(filtered_data_male$Total))
filtered_data_female <- filter(data, Sex == "Female", Age=="Total")
female = sum(as.numeric(filtered_data_female$Total))
data1 <- c(male,female)
barplot(data1, names.arg = c("Male","Female"), col = "skyblue", main = "Number of Suicides by Gender", xlab = "Sex", ylab = "Total")
```

```{r}
#| codefold: true
#| output: false
b1 <- filter(data,Sex=="Male",Age=="Total")
b2 <- filter(data,Sex=="Female",Age=="Total")

slices1 <- c(sum(as.numeric(b1$Total)),sum(as.numeric(b2$Total)))

lbls <- c("Male","Female")
pct <- round(slices1/sum(slices1)*100)
lbls <- paste(lbls, pct)
# add percents to labels
lbls <- paste(lbls,"%",sep="") # ad % to labels
```

```{r}
#| codefold: true
pie(slices1,labels = lbls, col=rainbow(length(lbls)),
   main="Pie Chart of Gender")
```

-From this graph, we can observe that **men are more prone to suicide compared to women**.

## **Suicides by Age Groups**

We used the bar plot method to compare suicides by age groups regardless of gender and reason. You can find the graph and the code we used below.

```{r}
#| code-fold: true
filtered_data_1 <- filter(data,Age=="<15", Sex == "Total")
filtered_data_2 <- filter(data,Age=="15-19", Sex == "Total")
filtered_data_3 <- filter(data,Age=="20-24", Sex == "Total")
filtered_data_4 <- filter(data,Age=="25-29", Sex == "Total")
filtered_data_5 <- filter(data,Age=="30-34", Sex == "Total")
filtered_data_6 <- filter(data,Age=="35-39", Sex == "Total")
filtered_data_7 <- filter(data,Age=="40-44", Sex == "Total")
filtered_data_8 <- filter(data,Age=="45-49", Sex == "Total")
filtered_data_9 <- filter(data,Age=="50-54", Sex == "Total")
filtered_data_10 <- filter(data,Age=="55-59", Sex == "Total")
filtered_data_11 <- filter(data,Age=="60-64", Sex == "Total")
filtered_data_12 <- filter(data,Age=="65-69", Sex == "Total")
filtered_data_13 <- filter(data,Age=="75 +", Sex == "Total")
age1 = sum(as.numeric(filtered_data_1$Total))
age2 = sum(as.numeric(filtered_data_2$Total))
age3 = sum(as.numeric(filtered_data_3$Total))
age4 = sum(as.numeric(filtered_data_4$Total))
age5 = sum(as.numeric(filtered_data_5$Total))
age6 = sum(as.numeric(filtered_data_6$Total))
age7 = sum(as.numeric(filtered_data_7$Total))
age8 = sum(as.numeric(filtered_data_8$Total))
age9 = sum(as.numeric(filtered_data_9$Total))
age10 = sum(as.numeric(filtered_data_10$Total))
age11 = sum(as.numeric(filtered_data_11$Total))
age12 = sum(as.numeric(filtered_data_12$Total))
age13 = sum(as.numeric(filtered_data_13$Total))
data2 = c(age1,age2,age3,age4,age5,age6,age7,age8,age9,age10,age11,age12,age13)
barplot(data2, names.arg = c("<15","15-19","20-24","25-29","30-34","35-39","40-44","45-49","50-54","55-59","60-64","65-69","75 +"), col = "red", main = "Number of Suicides by Age Group", xlab = "Age", ylab = "Total")
```

When looking at this graph, we observe that **suicides are most common among young adults** aged 20-24. We can say that as age progresses beyond 25, suicide rates tend to decrease. Another interesting result is that the age groups with the **lowest suicide rates are children and the elderly.**

## **Suicides by Year**

We used the bar plot method to compare suicides by year regardless of gender, reason and age group. You can find the graph and the code we used below.

```{r}
#| code-fold: true
#| output: false
ffiltered_data_6 <- filter(data,Year=="2007",Age=="Total", Sex == "Total")
ffiltered_data_7 <- filter(data,Year=="2008",Age=="Total", Sex == "Total")
ffiltered_data_8 <- filter(data,Year=="2009",Age=="Total", Sex == "Total")
ffiltered_data_9 <- filter(data,Year=="2010",Age=="Total", Sex == "Total")
ffiltered_data_10 <- filter(data,Year=="2011",Age=="Total", Sex == "Total")
ffiltered_data_11 <- filter(data,Year=="2012",Age=="Total", Sex == "Total")
ffiltered_data_12 <- filter(data,Year=="2013",Age=="Total", Sex == "Total")
ffiltered_data_13 <- filter(data,Year=="2014",Age=="Total", Sex == "Total")
ffiltered_data_14 <- filter(data,Year=="2015",Age=="Total", Sex == "Total")
ffiltered_data_15 <- filter(data,Year=="2016",Age=="Total", Sex == "Total")
ffiltered_data_16 <- filter(data,Year=="2017",Age=="Total", Sex == "Total")
ffiltered_data_17 <- filter(data,Year=="2018",Age=="Total", Sex == "Total")
ffiltered_data_18 <- filter(data,Year=="2019",Age=="Total", Sex == "Total")
ffiltered_data_19 <- filter(data,Year=="2020",Age=="Total", Sex == "Total")
ffiltered_data_20 <- filter(data,Year=="2021",Age=="Total", Sex == "Total")
ffiltered_data_21 <- filter(data,Year=="2022",Age=="Total", Sex == "Total")

year6 = sum(as.numeric(ffiltered_data_6$Total))
year7 = sum(as.numeric(ffiltered_data_7$Total))
year8 = sum(as.numeric(ffiltered_data_8$Total))
year9 = sum(as.numeric(ffiltered_data_9$Total))
year10 = sum(as.numeric(ffiltered_data_10$Total))
year11 = sum(as.numeric(ffiltered_data_11$Total))
year12 = sum(as.numeric(ffiltered_data_12$Total))
year13 = sum(as.numeric(ffiltered_data_13$Total))
year14 = sum(as.numeric(ffiltered_data_14$Total))
year15 = sum(as.numeric(ffiltered_data_15$Total))
year16 = sum(as.numeric(ffiltered_data_16$Total))
year17 = sum(as.numeric(ffiltered_data_17$Total))
year18 = sum(as.numeric(ffiltered_data_18$Total))
year19 = sum(as.numeric(ffiltered_data_19$Total))
year20 = sum(as.numeric(ffiltered_data_20$Total))
year21 = sum(as.numeric(ffiltered_data_21$Total))
```

```{r}
#| code-fold: true
data3 = c(year6,year7,year8,year9,year10,year11,year12,year13,year14,year15,year16,year17,year18,year19,year20,year21)
barplot(data3, names.arg = c("2007","2008","2009","2010","2011","2012","2013","2014","2015","2016","2017","2018","2019","2020","2021","2022"), col = "yellow", main = "Number of Suicides by Year", xlab = "Year", ylab = "Total")

```

When looking at this graph, we observe that **suicides are most common in 2021 and 2022**.

Factors such as people entering social isolation, losing their jobs, and experiencing financial difficulties due to the pandemic have made the phenomenon of suicide more painful.

## **Suicides by Reason**

We used the bar plot method to compare suicides by reason regardless of gender and age group. You can find the graph and the code we used below.

```{r}
#| code-fold: true
#| output: false
fffiltered_data_1 <- filter(data,Age =="Total", Sex == "Total")
cause1 = sum(as.numeric(fffiltered_data_1$Illness))
cause2 = sum(as.numeric(fffiltered_data_1$Family))
cause3 = sum(as.numeric(fffiltered_data_1$Economy))
cause4 = sum(as.numeric(fffiltered_data_1$Business))
cause5 = sum(as.numeric(fffiltered_data_1$Emotional))
cause6 = sum(as.numeric(fffiltered_data_1$Education))
cause7 = sum(as.numeric(fffiltered_data_1$Other))
unknown <- ifelse(is.na(as.numeric(fffiltered_data_1$Unknown)),0,as.numeric(fffiltered_data_1$Unknown))
cause8 = sum(unknown)
data4 = c(cause1,cause2,cause3,cause4,cause5,cause6,cause7,cause8)
```

```{r}
#| code-fold: true
barplot(data4, names.arg = c("Illness","Family","Economy","Business","Emotional","Educational","Other","Unknown"), col = "brown", main = "Number of Suicides by Reason", xlab = "Suicide Reason", ylab = "Total",cex.names=0.72)
```

This line chart shows the change in the causes of suicide over the years.

```{r}
#| codefold: true
#| output: false
# Keep only 3 names
  a = filter(data,Age=="Total", Sex =="Total")
  a1 = data.frame(Year=a$Year,Total=a$Illness,Cause="Illness")
  a2 = data.frame(Year=a$Year,Total=a$Family,Cause="Family")
  a3 = data.frame(Year=a$Year,Total=a$Economy,Cause="Economy")
  a4 = data.frame(Year=a$Year,Total=a$Business,Cause="Business")
  a5 = data.frame(Year=a$Year,Total=a$Emotional,Cause="Emotional")
  a6 = data.frame(Year=a$Year,Total=a$Education,Cause="Education")
  a7 = data.frame(Year=a$Year,Total=a$Other,Cause="Other")
  a8 = data.frame(Year=a$Year,Total=a$Unknown,Cause="Unknown")
  cause = rbind(a1,a2,a3,a4,a5,a6,a7,a8,deparse.level = 0)
```

```{r}
#| codefold: true
graph <- cause %>% 
    filter(Cause %in% c("Illness","Family","Economy","Business","Emotional","Education","Other","Unknown"))
  b = as.numeric(cause$Total)
  c = ifelse(is.na(b),0,b)
  graph %>%
    ggplot( aes(x=Year, y=c, group=Cause, color=Cause)) +
      geom_line() +
      ylab("Number of Suicides") +
      ggtitle("Number of Suicides by Reasons")
```

We can say that while the number of suicides due to business and education has decreased over time, the number of other suicides has remained almost constant.

We used a pie chart to better analyze the ratios.

```{r}
# Pie Chart with Percentages
#| codefold: true
#| output: false
slices <- c(
  sum(ifelse(is.na(as.numeric(a$Illness)),0,as.numeric(a$Illness))),
  sum(ifelse(is.na(as.numeric(a$Family)),0,as.numeric(a$Family))) ,
  sum(ifelse(is.na(as.numeric(a$Economy)),0,as.numeric(a$Economy))),
  sum(ifelse(is.na(as.numeric(a$Business)),0,as.numeric(a$Business))),
  sum(ifelse(is.na(as.numeric(a$Emotional)),0,as.numeric(a$Emotional))),
  sum(ifelse(is.na(as.numeric(a$Education)),0,as.numeric(a$Education))),
  sum(ifelse(is.na(as.numeric(a$Other)),0,as.numeric(a$Other))),sum(ifelse(is.na(as.numeric(a$Unknown)),0,as.numeric(a$Unknown))))

lbls <- c("Illness","Family","Economy","Business","Emotional","Education","Other","Unknown")
pct <- round(slices/sum(slices)*100)
lbls <- paste(lbls, pct)
# add percents to labels
lbls <- paste(lbls,"%",sep="") # ad % to labels
```

```{r}
#| codefold: true
pie(slices,lbls, col=rainbow(length(lbls)),
   main="Pie Chart of Suicide Reasons",cex=0.8)
```

As a result, we see that if we ignore the unknowns and others, the Illness has a huge dominance.

### **Suicides of Males by Reason**

To obtain more meaningful results, we separated the data by gender.

We interpreted the reasons why men commit suicide with a line chart.

```{r}
#| codefold: true
#| output: false
  a = filter(data,Age=="Total", Sex == "Male")
  a1 = data.frame(Year=a$Year,Total=a$Illness,Cause="Illness")
  a2 = data.frame(Year=a$Year,Total=a$Family,Cause="Family")
  a3 = data.frame(Year=a$Year,Total=a$Economy,Cause="Economy")
  a4 = data.frame(Year=a$Year,Total=a$Business,Cause="Business")
  a5 = data.frame(Year=a$Year,Total=a$Emotional,Cause="Emotional")
  a6 = data.frame(Year=a$Year,Total=a$Education,Cause="Education")
  a7 = data.frame(Year=a$Year,Total=a$Other,Cause="Other")
  a8 = data.frame(Year=a$Year,Total=a$Unknown,Cause="Unknown")
  cause = rbind(a1,a2,a3,a4,a5,a6,a7,a8,deparse.level = 0)
  b = as.numeric(cause$Total)
  c = ifelse(is.na(b),0,b)
```

```{r}
#| codefold: true
graph <- cause %>% 
    filter(Cause %in% c("Illness","Family","Economy","Business","Emotional","Education","Other","Unknown"))
  graph %>%
    ggplot( aes(x=Year, y=c, group=Cause, color=Cause)) +
      geom_line() +
      ylab("Number of Suicides") +
      ggtitle("Number of Males' Suicides by Reasons ")
```

```{r}
#| codefold: true
#| output: false
slices2 = c(sum(as.numeric(a1$Total)),sum(as.numeric(a2$Total)),sum(as.numeric(a3$Total)),sum(as.numeric(a4$Total)),sum(as.numeric(a5$Total)),sum(ifelse(is.na(as.numeric(a6$Total)),0,as.numeric(a6$Total))),sum(as.numeric(a7$Total)),sum(as.numeric(a8$Total)))
name = c("Illness","Family","Economy","Business","Emotional","Education","Other","Unknown")
lbls2 <- name
pct <- round(slices2/sum(slices2)*100)
lbls2 <- paste(lbls2, pct)
# add percents to labels
lbls2 <- paste(lbls2,"%",sep="") # ad % to labels
```

```{r}
#| codefold: true

pie(slices2,lbls2, col=rainbow(length(lbls2)),
   main="Pie Chart of Males' Suicide by Reasons",cex=0.8)

```

### **Suicides of Females by Reason**

```{r}
#| codefold: true
#| output: false
  a = filter(data,Age=="Total", Sex == "Female")
  a1 = data.frame(Year=a$Year,Total=a$Illness,Cause="Illness")
  a2 = data.frame(Year=a$Year,Total=a$Family,Cause="Family")
  a3 = data.frame(Year=a$Year,Total=a$Economy,Cause="Economy")
  a4 = data.frame(Year=a$Year,Total=a$Business,Cause="Business")
  a5 = data.frame(Year=a$Year,Total=a$Emotional,Cause="Emotional")
  a6 = data.frame(Year=a$Year,Total=a$Education,Cause="Education")
  a7 = data.frame(Year=a$Year,Total=a$Other,Cause="Other")
  a8 = data.frame(Year=a$Year,Total=a$Unknown,Cause="Unknown")
  cause = rbind(a1,a2,a3,a4,a5,a6,a7,a8,deparse.level = 0)
  b = as.numeric(cause$Total)
  c = ifelse(is.na(b),0,b)
```

```{r}
#| codefold: true
graph <- cause %>% 
    filter(Cause %in% c("Illness","Family","Economy","Business","Emotional","Education","Other","Unknown"))
  graph %>%
    ggplot( aes(x=Year, y=c, group=Cause, color=Cause)) +
      geom_line() +
      ylab("Number of Suicides") +
      ggtitle("Number of Females' Suicide by Reasons")
```

```{r}
#| codefold: true
#| output: false
slices = c(sum(as.numeric(a1$Total)),sum(as.numeric(a2$Total)),sum(as.numeric(a3$Total)),sum(ifelse(is.na(as.numeric(a4$Total)),0,as.numeric(a4$Total))),sum(ifelse(is.na(as.numeric(a5$Total)),0,as.numeric(a5$Total))),sum(ifelse(is.na(as.numeric(a6$Total)),0,as.numeric(a6$Total))),sum(as.numeric(a7$Total)),sum(as.numeric(a8$Total)))
name = c("Illness","Family","Economy","Business","Emotional","Education","Other","Unknown")
lbls <- name
pct <- round(slices/sum(slices)*100)
lbls <- paste(lbls, pct)
# add percents to labels
lbls <- paste(lbls,"%",sep="") # ad % to labels
```

```{r}
#| codefold: true
pie(slices,labels = lbls, col=rainbow(length(lbls)),
   main="Pie Chart of Females' Suicide by Reasons",cex=0.5)
```
